---
title: "Assignment 2: Coding Basics"
author: "Kathlyn MacDonald"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. 
#Creating a sequence of #s from 1 to 100, going up by 4s; and assigning to the name sequence_by4:
sequence_by4 <- seq(1,100,4)
#calling the sequence to the console:
sequence_by4

#2. 
#Calculate the average of sequence_by4
mean(sequence_by4)
#Find the middle value in sequence_by4
median(sequence_by4)

#3.
#Asking if the mean is greater than the median:
mean(sequence_by4)>median(sequence_by4)


```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}

#Vector character list with all names of the students
studentNames<- c('John','Emily',"Mary","Jodie","Sunny","Jackie","Atalie","Jimmy","Jiblet","Karen","Samantha","Jessica","Angela","Dwight","Syd","Mrs.Poppins","Janet","Ted","Mr.Mosby","Vincent","Arthur","Candy","Junebug","Emmy","Roger")

#Numerical list with all the grades the students received

grades<-c(75,56,90,48,67,89,35,56,64,73,45,97,84,35,25,55,86,75,90,90,43,76,41,84,79)

#If else statement assigning fail if the grade is less then 50, and pass if the grade is above 50
TrueOrFalse<- ifelse(grades<50,"False","True")

TrueOrFalse

#Creating the data frame
df_names <- as.data.frame(studentNames)
#Binding the grades and pass/fail status to the data frame
df_StudentList <-cbind(df_names,grades,TrueOrFalse)

class(df_StudentList)

```

9. QUESTION: How is this data frame different from a matrix?

> Answer: A matrix must have elements of all the same type, this data frame contains both numerical values and characters.

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
#Generating pass or fail status, depending on if the grade is above 50
PassFailStatus<-
  ifelse(TrueOrFalse=="True",print("Pass"),print("Fail"))
PassFailStatus


#
df_StudentList <-cbind(df_names,grades,TrueOrFalse,PassFailStatus)
df_StudentList

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: I only got the ifelse statement to work. I got an error message with the if else statement, that the condition was too long, and it wouldn't check all of the elements. I would assume that means that the if else function doesn't work with a vector, and only works wth one element.
