---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Kathlyn MacDonald"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Monday, February 28 at 7:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1

getwd()
library(tidyverse)
library(agricolae)

NTL_LTER <- read.csv("../Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)

NTL_LTER$sampledate <- as.Date(NTL_LTER$sampledate, format = "%m/%d/%y")

#2

mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature is not related to depth across all lakes, when tested in July.
Ha: Mean lake temperature and depth are related across all lakes, when tested in July.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

NTL_LTER_wrangled <-
  NTL_LTER %>%
  filter(months(NTL_LTER$sampledate)== "July") %>%
  select(`lakename`, `year4`, `daynum`, `depth`, `temperature_C`) %>%
  drop_na()
  


#5

Temperature_Depth <- ggplot(NTL_LTER_wrangled, aes(x= temperature_C, y = depth)) +
  geom_point() +
  geom_smooth(method = "lm", color = 'red')+
  labs(title = "July lake temperature at varying depth", y = "Depth (m)", x = "Temperature (C)")+
  xlim(0, 35) +
  scale_y_reverse()
print(Temperature_Depth)
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: This graph suggests that as depth increases, temperature decreases. The point distribution is fairly constant indicating the results are robust and reliable (over the course of many years, each year saw similar results). 


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
temp_depth.regression <- lm(data = NTL_LTER_wrangled, formula = temperature_C ~ depth)

summary(temp_depth.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The estimated effect of depth on temperature is -1.95. Which means that for every 1m increase in depth, there is a 1.95C drop in temperature. The R squared value is high, at 0.7387, indicating that depth explains 73.87% of the variability. There is a relatively small standard error, indicating accurate predictions. This result is based off of 9726 degrees of freedom. The results here are significant, so we can reject the null hypothesis. Therefore, there is a relationship between temperature and depth.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
TPAIC <- lm(data = NTL_LTER_wrangled, temperature_C ~ depth + daynum + year4)

step(TPAIC)

#10
Temp_multiple.regression <- lm(data = NTL_LTER_wrangled, temperature_C ~ depth + daynum + year4)
summary(Temp_multiple.regression)


```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC model recommended that we use all 3 variables, depth, daynum and year4. 74.12% of variability in temperature, is explained by these three variables. Which is only a very slight improvement from the model using only depth as a variable (73%).



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

Lake.temp.anova <- aov(data = NTL_LTER_wrangled, temperature_C ~ lakename)
summary(Lake.temp.anova)

Lake.temp.anova2 <- lm(data = NTL_LTER_wrangled, temperature_C ~ lakename)
summary(Lake.temp.anova2)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: There is a significant difference, so we can reject the null. The means are not all the same across all lake sites.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

Temp_depth_lakename <- ggplot(NTL_LTER_wrangled, aes(x = depth, y = temperature_C, color = lakename))+
  geom_point(alpha = 0.5)+
  geom_smooth(method = "lm", se = FALSE)+
  xlim(0,15) +
  ylim(0,35) +
  labs (title = "Lake Temperature by Depth", x = "Depth (m)", y = "Temperature (C)") + 
  theme(legend.position = "bottom")
print(Temp_depth_lakename)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(Lake.temp.anova)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: The lakes that have the same mean temperature as Peter lake are Paul Lake and Ward Lake (P> 0.05, therefore we cannot reject the null that there is no difference between means).No they all have at least one other lake that they share the same statistically mean lake temperature. 



 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: Pairwise interaction test between the two variables, and then an anova and tukey HSD test to determine significance


